---
title: "Advent of Code 2023: Day 1"
output: html_notebook
---

# Part One

Start by loading in our data:

```{r}
#https://adventofcode.com/2023/day/1
day1 <- read.table(file = paste0(getwd(), "/data/day1.txt"))
```

We want to concatenate the first- and last-observed numerics in each string to end up with a single number per row, then sum all those numbers together.

Per the example, there's an edge case to consider: a row string may only contain a single numeric. Not a big deal.

I'll start with the easiest (read: dumbest) solution I can think of:

```{r}
library(stringr) #for str_split function
numbers <- sapply(day1$V1
                  , FUN = function(x){
  tmp <- suppressWarnings(stringr::str_split_1(x, "") %>%
    as.numeric(.) %>% #force all vector elements to numeric
    na.omit(.)) #drop all NAs
  
  #should be left with a vector of numerics now
  #concatenate the first and last elements of tmp, 
  #then force to numeric
  as.numeric(paste0(tmp[1], tail(tmp,1)))
}
, USE.NAMES = FALSE)
```

Now we have a vector of numbers and sum them up:

```{r}
sum(numbers)
```

And this turned out to be the right answer. I'll likely come back later to see if I can come up with a more elegant solution.

# Part Two

Trickier! We could build a look-up table and search each row string for matches? Still need to account for order, which makes the string matching a bit more nuanced.

```{r}
written.nums <- c("one"
                  , "two"
                  , "three"
                  , "four"
                  , "five"
                  , "six"
                  , "seven"
                  , "eight"
                  , "nine")
```

Let's play with an example or two from the prompt. The second one here illustrates one of the hang-ups: some spelled-out numbers may share characters (e.g. `twone`)

```{r}
ex1 <- "two1nine"
ex2 <- "xtwone3four"
```

Might be easiest to use `gsub`, although my solution will be clunky since I don't know regex:

```{r}

ex1.chk <- mapply(written.nums
       , 1:9
       , ex1
       , FUN = function(x, y, z){
         gsub(x,y,z)
       }
       , USE.NAMES = FALSE)

ex1.chk
```

We end up with a string vector we could take advantage of: drop all elements that match the original string, then pluck the first and last elements from the remainder to obtain the string-to-numeric values for downstream use.

```{r}
ex1.chk[ex1.chk != ex1]
```

In this case there are only two elements; seems like we can use a recursion on these elements to further simplify things, stopping when the only elements that remain are identical:

```{r}
#define helper
`%notin%` <- Negate(`%in%`)

ex1.chk2 <- suppressWarnings(
  mapply(written.nums
         , 1:9
         , ex1.chk[ex1.chk != ex1]
         , FUN = function(x, y, z){
           gsub(x,y,z)
         }
         , USE.NAMES = FALSE)
)

ex1.result <- ex1.chk2[ex1.chk2 %notin% ex1.chk[ex1.chk != ex1]]
ex1.result[c(1, length(ex1.result))]

```

And from here we pluck the first and last characters from the remaining element, then concatenate.

Let's try the other example:

```{r}

```

