---
title: 'Advent of Code 2023: Day 1'
output:
  html_document:
    df_print: paged
---

```{r setup, echo = FALSE}

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

```

# Part One

Start by loading in our data:

```{r}

day1 <- read.table(file = paste0(getwd(), "/data/day1.txt"))
```

We want to concatenate the first- and last-observed numerics in each string to end up with a single number per row, then sum all those numbers together.

Per the example, there's an edge case to consider: a row string may only contain a single numeric. Not a big deal.

## Proposed Solution

I'll start with the easiest (read: dumbest) solution I can think of:

```{r}
library(stringr) #for str_split function
numbers <- sapply(day1$V1
                  , FUN = function(x){
  tmp <- suppressWarnings(stringr::str_split_1(x, "") %>%
    as.numeric(.) %>% #force all vector elements to numeric
    na.omit(.)) #drop all NAs
  
  #should be left with a vector of numerics now
  #concatenate the first and last elements of tmp, 
  #then force to numeric
  as.numeric(paste0(tmp[1], tail(tmp,1)))
}
, USE.NAMES = FALSE)
```

Now we have a vector of numbers and sum them up:

```{r}
sum(numbers)
```

And this turned out to be the right answer. I'll likely come back to this one later to see if I can come up with a more elegant solution.

# Part Two

Trickier! We could build a look-up table and search each row string for matches? Still need to account for order, which makes the string matching a bit more nuanced.

```{r}
written.nums <- c("one"
                  , "two"
                  , "three"
                  , "four"
                  , "five"
                  , "six"
                  , "seven"
                  , "eight"
                  , "nine")
```

## Playing with the Problem

Let's play with an example or two from the prompt. The second one here illustrates one of the hang-ups: some spelled-out numbers may share characters (e.g. `twone`)

```{r}
ex1 <- "two1nine"
ex2 <- "xtwone3four"
```

Might be easiest to use `gsub`, although my solution will be clunky since I don't know regex:

```{r}

ex1.chk <- mapply(written.nums
       , 1:9
       , ex1
       , FUN = function(x, y, z){
         gsub(x,y,z)
       }
       , USE.NAMES = FALSE)

ex1.chk
```

We end up with a string vector we could take advantage of: drop all elements that match the original string, then pluck the modified string with the earliest deviation in "spelling" from the original string.

```{r}
ex1.chk[ex1.chk != ex1]
```

In this case we'd pluck the first vector element, then pass over the string and determine whether there's some last-observed word that needs replacing with its numeric symbol.

## Proposed Solution

Let's implement this approach in some defined functions and then test it on the example vector of strings from the part two problem statement:

```{r}
#helper
`%notin%` <- Negate(`%in%`)

wordReplace <- function(strng, reverse = FALSE){
  
  #take a first pass at replacing all possible words with their numeric symbol equivalent
  word.list <- 
    sapply(strng, function(a) {mapply(written.nums
                                      , 1:9
                                      , a
                                      , FUN = function(x, y, z){
                                        gsub(x,y,z)
                                      }
                                      , USE.NAMES = FALSE)}
           , simplify = FALSE
    ) %>% 
    unlist(.) %>% 
    unique(.) #retain unique strings only
  
  if (length(unique(word.list)) == 1){
    #nothing else to do
    return(unique(word.list))
  } else {
    #compare each of the modified strings with the original string,
    #then determine the one with earliest deviation vs. the original
    keep <- sapply(word.list[word.list %notin% strng]
                   , FUN = function(x){
                     
                     val1 <- str_split(x, "", simplify = TRUE)
                     val2 <- str_split(strng, "", simplify = TRUE)
                     
                     #use reverse = TRUE to ID the modified string with the last possible word replaced with its numeric symbol equivalent
                     if (reverse){
                       val1 <- rev(val1)
                       val2 <- rev(val2)
                     }
                     
                     #find the index of the modified string element that first deviates from that of the original string
                     chk.min <- if(length(val1) < length(val2)){
                       which.min(val1 == val2[1:length(val1)])
                     } else {
                       which.min(val1[1:length(val2)] == val2)
                     }
                     
                     chk.min
                   }
                   , simplify = TRUE)
    
    #return the modified string with the earliest deviation from the original string
    return(names(keep)[which.min(keep)])
  }
}

batchReplace <- function(strngs){
  #run the wordReplace routine on each string in strngs
  sapply(strngs
         , FUN = function(x){
           init.word <- wordReplace(x)
           wordReplace(init.word, reverse = TRUE)
         })
}

```

Let's also define a function to handle the integer-search routine from before.

```{r}
#ID the first and last numerics for each element in a vector of strings,
#then concatenate them
numberSearch <- function(string.vec){
  if (any(is.na(string.vec))) browser()
  sapply(string.vec
         , FUN = function(x){
           tmp <- suppressWarnings(stringr::str_split_1(x, "") %>%
                                     as.numeric(.) %>% #force all vector elements to numeric
                                     na.omit(.)) #drop all NAs
           
           #should be left with a vector of numerics now
           #concatenate the first and last elements of tmp, 
           #then force to numeric
           as.numeric(paste0(tmp[1], tail(tmp,1)))
         }
  )
}
```

Next test these functions on the example data, since we know what the result should be:

```{r}
day1.test <- read.table(file = paste0(getwd(), "/data/day1_sample.txt"))

day1.test.str <- sapply(day1.test$V1
                           , batchReplace
                           , USE.NAMES = TRUE)

sum(numberSearch(unlist(day1.test.str)))
```

Looks like we're in good shape. Now let's run these functions on the full dataset:

```{r}

#string-and-replace first
day1.str.replace <- sapply(day1$V1
                           , batchReplace
                           , USE.NAMES = TRUE)

#number search second
day1.final.pt2 <- numberSearch(unlist(day1.str.replace))

sum(day1.final.pt2)
```

Turns out this answer is wrong! There are some strings where words share a common letter, e.g. `twone`. I assumed I would only need to account for the first word that appears in such a phrase (`two`), but maybe I need to account for both (`two`, `one`).

## Proposed Solution No. 2

Back to the drawing board. I think we can slightly modify the functions from the previous section to catch these edge cases.

```{r}
#ID the first numeric for each element in a string
numberSearch2 <- function(strng, reverse = FALSE){
  tmp <- if (reverse){
    suppressWarnings(stringr::str_split_1(strng, "") %>%
                       rev(.) %>%
                           as.numeric(.) %>% #force all vector elements to numeric
                           na.omit(.)) #drop all NAs
  } else {
    suppressWarnings(stringr::str_split_1(strng, "") %>%
                           as.numeric(.) %>% #force all vector elements to numeric
                           na.omit(.)) #drop all NAs
  }
  #should be left with a vector of numerics
  tmp[1]
}

batchReplace2 <- function(strngs){
  #run the wordReplace routine on each string in strngs
  sapply(strngs
         , FUN = function(x){
           init.word <- wordReplace(x)
           init.number <- numberSearch2(init.word)
           last.word <- wordReplace(x, reverse = TRUE)
           last.number <- numberSearch2(last.word, reverse = TRUE)
           as.numeric(paste0(init.number,last.number))
         })
}
```

Test again on the example:

```{r}
day1.test2 <- sapply(day1.test$V1
                           , batchReplace2
                           , USE.NAMES = TRUE)

sum(day1.test2)
```

Looks like we're still in good shape. Let's also run it on one of the edge cases:

```{r}
day1.test3 <- sapply("twone"
                           , batchReplace2
                           , USE.NAMES = TRUE)

sum(day1.test3)
```

Looking good. Now let's run these functions on the full dataset:

```{r}

#string-and-replace first
day1.str.replace2 <- sapply(day1$V1
                           , batchReplace2
                           , USE.NAMES = TRUE)

sum(day1.str.replace2)
```

Turns out this answer is correct.